---
title: "shippping random walk"
# subtitle: ""
# bibliography: linen_quality.bib
format: 
  html:
    mainfont: Quattrocento Sans
    sansfont: Cinzel
    toc: true
    toc-location: left
    toc-depth: 3
    fig-cap-location: margin
execute: 
  echo: false
  message: false
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(ggtext)
library(here)
library(rstan)

here::i_am("ship_random_walk.qmd")

# stan options
util <- new.env()
rstan_options(auto_write = TRUE)
source(here("mcmc_analysis_tools_rstan.R"), local = util)
source(here("mcmc_visualization_tools.R"), local = util)

# graphics options
extrafont::font_import(pattern = "QuattrocentoSans", prompt = FALSE)
extrafont::loadfonts(device = "win")
ggplot2::theme_set(theme_minimal() +
                     theme(text = element_text(size = 15,
                                               family = "Quattrocento Sans"),
                           margins = margin(10, 15, 25, 15),
                           axis.title.x = element_text(vjust = -3),
                           axis.title.y = element_text(vjust =  3)))
options(knitr.table.html.attr = "quarto-disable-processing=true")

# inline content
inline <- list()

c_light <- c("#D6EBEA")
c_light_highlight <- c("#B5DBD9")
c_mid <- c("#92CAC6")
c_mid_highlight <- c("#5FB2AC")
c_dark <- c("#369F97")
c_dark_highlight <- c("#00867B")

c_dark_trans <- c("#005D5580")
c_purple_trans <- c("#D400C380")

c_list <- list(c_light, c_light_highlight, c_mid, 
                 c_mid_highlight, c_dark, c_dark_highlight)
```

# priors

shipping delay should be bounded below at 0 with a long upper tail, 80% should be between say 0 and 15 days with median maybe 3. this is location dependent and we have location data so maybe we'll work that in later

the production rate is specified here in terms of 'days per game' i think we need to be careful about this parameterization. our domain knowledge is in 'games per week', we might expect the 10/50/90 percentiles to be 5, 10, 20. That's about 0.3, 0.7, and 1.4 days per game. But as it gets below 0.3 the number of games per week becomes implausible, for example 0.1 = 70 games per week, so we want to really keep it far from 0. inverse gamma seems like a great idea

# simulated dataset

```{r}
ex_data <- 
  tibble(
    relic = 1:720
  ) %>%
  mutate(prod_delay = 20,
         prod_error = extraDistr::rinvgamma(n = n(), 2, 1),
         prod = prod_delay + cumsum(prod_error),
         ship_error = rexp(n = n(), rate = 1/7),
         # ship_error = rexp(n = n(), rate = 1/5),
         deliv = prod + ship_error,
         is_observed = row_number() %in% sample(1:650, size = 300, replace = FALSE))

ex_data %>%
  ggplot(aes(x = relic)) +
  geom_point(data = . %>% filter(!is_observed),
             aes(y = deliv, color = is_observed)) +
  geom_line(data = . %>% filter(relic <= 650),
            aes(y = prod)) +
  geom_point(data = . %>% filter(is_observed),
             aes(y = deliv, color = is_observed)) +
  scale_color_manual(guide = "none",
                     values = c("TRUE"= "#000000", "FALSE" = "#CCCCCC"))
```

::: {.callout-important}
write observational model down
:::

# prior predictive

```{r}
writeLines(readLines(here("stan_programs/ship_random_walk_bayesian_ensemble.stan")))
```

```{r}
fit_ensemble <- stan(
  file = here("stan_programs/ship_random_walk_bayesian_ensemble.stan"),
  data = list(N = nrow(ex_data)),
  iter = 5000, chains = 1,
  warmup = 0, algorithm = "Fixed_param",
  refresh = 1000
)
```

```{r}
# idx <- rep(1:nrow(ex_data), each = 2)
# x <- sapply(X = 1:length(idx), 
#             FUN = \(b){if(b %% 2 == 0) {idx[b] + 0.5} else {idx[b] - 0.5}})
# cred <- sapply(1:nrow(ex_data), \(i){quantile(extract(fit_ensemble)$y[ , i], probs = seq(0.1, 0.9, by = 0.1))})
n <- ncol(extract(fit_ensemble)$production)

plot(1, type = "n", 
     xlim = c(1, n), ylim = c(0, 300/0.5),
     xlab = "y", main = "Prior Predictive")
# for (i in 1:((nrow(cred) - 1)/2)) {
#   polygon(x = c(x, rev(x)),
#           y = c(rep(cred[i, ], each = 2), 
#                 rep(rev(cred[nrow(cred) - i + 1, ]), each = 2)),
#           col = c_list[[i]], border = NA)
# }
# lines(x, rep(cred[5, ], each = 2), 
#       col = c_dark, lwd = 3)
for (i in 1:20) {
  lines(x = 1:n,
        y = extract(fit_ensemble)$production[i, ],
        col = c_dark_trans, lwd = 2)
}
lines(x = 1:n,
      y = 20 + 1:n * 1/0.7)
```

# posterior retrodictive

```{r}
writeLines(readLines(here("stan_programs/ship_random_walk.stan")))
```

it works well for low volumes of simulated data but barfs when you get too many observations
that's the only insight i have right now

```{r}
stan_data <- list(
  N = nrow(ex_data),
  N_obs = sum(ex_data$is_observed),
  relic_obs = ex_data$relic[ex_data$is_observed],
  relic_unobs = ex_data$relic[!ex_data$is_observed],
  delivery_obs = ex_data$deliv[ex_data$is_observed]
)

fit <- stan(
  file = here("stan_programs/ship_random_walk.stan"),
  data = stan_data, 
  # init = \(){list(prod_delay = 20, prod_mu = 1, prod_sigma = 0.2, ship_lambda = 1/3,
  #                 delivery_unobs = stan_data$relic_unobs * 0.5,
  #                 production = (1:stan_data$N) * 0.5,
  #                 shipping = (1:stan_data$N) * 3)},
  iter = 2000, chains = 1, cores = 1, refresh = 1000, warmup = 500,
  control = list(adapt_delta = 0.9, max_treedepth = 15)
)
```

```{r}
{par(mfrow = c(2, 2))
hist(extract(fit)$prod_delay)
hist(extract(fit)$prod_mu)
hist(extract(fit)$prod_sigma)
hist(extract(fit)$ship_lambda)
par(mfrow = c(1, 1))}

as_tibble(extract(fit)) %>%
  select(starts_with(c("prod_", "ship_"))) %>%
  mutate(across(.cols = everything(),
                .fns  = log)) %>%
  pairs()

# prior contraction
# <go do betancourt's contraction thing>
```

```{r}
idx <- rep(1:stan_data$N, each = 2)
x <- sapply(X = 1:length(idx), 
            FUN = \(b){if(b %% 2 == 0) {idx[b] + 0.5} else {idx[b] - 0.5}})
cred <- sapply(1:stan_data$N, \(i){quantile(extract(fit)$production[ , i], probs = seq(0.1, 0.9, by = 0.1))})

plot(1, type = "n", 
     xlim = c(0, max(x)), ylim = c(0, max(cred)),
     xlab = "y", main = "Posterior Retrodictive")
for (i in 1:((nrow(cred) - 1)/2)) {
  polygon(x = c(x, rev(x)),
          y = c(rep(cred[i, ], each = 2), 
                rep(rev(cred[nrow(cred) - i + 1, ]), each = 2)),
          col = c_list[[i]], border = NA)
}
lines(x, rep(cred[5, ], each = 2), 
      col = c_dark, lwd = 1)
points(x = stan_data$relic_obs, y = stan_data$delivery,
       col = c_purple_trans, pch = 19)
# lines(x, rep(stan_data$y, each = 2), 
#       col = c_purple_trans, lwd = 3)
```

there's no latent measurement here like for rounding error so no more interesting summaries (i don't think?)

# Dune

use it on dune

```{r}
delivery_raw <- read_csv(here("data/raw/delivery_data.csv"))

dune <- 
  delivery_raw %>%
  filter(game == "Dune", 
         !is.na(relic),
         !is.na(delivered),
         !(relic %in% c(756, 711, 624))) %>%
  mutate(delivered = as.integer(delivered - as.Date("2025-04-14"))) %>%
  arrange(relic)
```

```{r}
a <- 
  lm(formula = delivered ~ relic,
     data = dune) %>%
  residuals() %>%
  `-`(., min(.))

plot(sort(a), extraDistr::qhnorm(ppoints(a), 4))
```

```{r}
# trying it with truncated real data
stan_data_dune <- list(
  N = 200,
  N_obs = nrow(dune[dune$relic <= 200, ]),
  relic_obs = dune$relic[dune$relic <= 200],
  relic_unobs = setdiff(1:200, dune$relic[dune$relic <= 200]),
  delivery_obs = dune$delivered[dune$relic <= 200]
)

plot(x = 1, type = "n", 
     xlim = c(0, max(stan_data_dune$relic_obs)),
     ylim = c(0, max(stan_data_dune$delivery)))
points(x = stan_data_dune$relic_obs, 
       y = stan_data_dune$delivery,
       col = c_purple_trans, pch = 19)

fit_dune <- stan(
  file = here("stan_programs/ship_random_walk.stan"),
  data = stan_data_dune,
  iter = 1000, chains = 1, cores = 1, refresh = 1000
)
```

something's really wrong, are the chains not mixing?

```{r}
hist(extract(fit_dune)$prod_delay)
hist(extract(fit_dune)$prod_mu)
hist(extract(fit_dune)$prod_sigma)
hist(extract(fit_dune)$ship_sigma)
hist(extract(fit_dune)$ship_lambda)

as_tibble(extract(fit_dune)) %>%
  # sample_n(5000) %>%
  select(starts_with(c("prod_", "ship_"))) %>%
  mutate(across(.cols = everything(),
                .fns  = log)) %>%
  pairs()
```

```{r}
idx <- rep(1:stan_data_dune$N, each = 2)
x <- sapply(X = 1:length(idx), 
            FUN = \(b){if(b %% 2 == 0) {idx[b] + 0.5} else {idx[b] - 0.5}})
cred <- sapply(1:stan_data_dune$N, \(i){quantile(extract(fit_dune)$production[ , i], probs = seq(0.1, 0.9, by = 0.1))})

plot(x = 1, type = "n", 
     xlim = c(0, max(x)), 
     ylim = c(0, max(cred)),
     # ylim = c(min(cred, stan_data_dune$delivery_obs),
     #          max(cred, stan_data_dune$delivery_obs)),
     xlab = "y", main = "Posterior Retrodictive")
for (i in 1:((nrow(cred) - 1)/2)) {
  polygon(x = c(x, rev(x)),
          y = c(rep(cred[i, ], each = 2), 
                rep(rev(cred[nrow(cred) - i + 1, ]), each = 2)),
          col = c_list[[i]], border = NA)
}
lines(x, rep(cred[5, ], each = 2), 
      col = c_dark, lwd = 1)
points(x = stan_data_dune$relic_obs, 
       y = stan_data_dune$delivery,
       col = c_purple_trans, pch = 19)

# plot(x = 1, type = "n", 
#      xlim = c(0, max(stan_data_dune$relic_obs)),
#      ylim = c(0, max(stan_data_dune$delivery)))
# points(x = stan_data_dune$relic_obs, 
#        y = stan_data_dune$delivery,
#        col = c_purple_trans, pch = 19)
```
